---
title: "Bioinformatics workshop scRNA-seq DE"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5,
  fig.align = "center"
)
```

# Setup

## Install packages

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install(c("muscat", "DESeq2", "SingleCellExperiment", "scater", "edgeR"))
install.packages(c("ggplot2", "dplyr", "pheatmap", "ggrepel"))
```

## Load libraries

```{r libraries}
library(muscat)
library(ExperimentHub)
library(SingleCellExperiment)
library(scater)
library(edgeR)
library(ggplot2)
library(dplyr)
library(pheatmap)
library(ggrepel)
```

# Load example data

```{r load-data}
# ExperimentHub is a bioconductor service that hosts and distributes curated datasets
# Can be queried by accession IDs
eh <- ExperimentHub()
sce <- eh[["EH2259"]]

# Inspect the object
sce
```

```{r explore-data}
# Available metadata
head(colData(sce))

# cell — cell type label (e.g., "B cells", "CD14+ Monocytes")
# ind — donor/individual ID (e.g., 107, 1016, 1256)
# stim — condition ("ctrl" or "stim")
# cluster — numeric cluster ID from the original clustering
# multiplets — whether the cell is a "singlet" or "doublet" (two cells captured together)

# Cell type labels
table(sce$cell)

# Sample IDs
table(sce$ind)

# Condition (group)
table(sce$stim)

# Cross-tabulation: cells per sample × cell type
table(sce$cell, sce$ind)
```

# Step 1: Prepare the SCE object

```{r prep-sce}
# Create unique sample ID per donor × condition
sce$sample <- paste0(sce$ind, "_", sce$stim)
# The EH2259 data already uses muscat's standard column names,
# but we still run prepSCE to set up internal metadata (experiment_info)
sce <- prepSCE(sce,
  kid = "cell", #your cluster column cluster_id (e.g., from Seurat clustering or manual annotation)
  sid = "sample", #your sample column sample_id (biological replicate / donor)
  gid = "stim") #your group column group_id (e.g., "healthy" vs. "disease")
sce <- logNormCounts(sce)
head(colData(sce)[, c("cluster_id", "sample_id", "group_id")])
```

```{r prep-summary}
# Quick summary of the experimental design
nlevels(sce$cluster_id)  # number of cell types
nlevels(sce$sample_id)   # number of samples
nlevels(sce$group_id)    # number of conditions
levels(sce$group_id)     # condition labels

# Cells per cluster × sample
table(sce$cluster_id, sce$sample_id)
```

# Step 2: Aggregate to pseudobulk

```{r aggregate}
# sums the raw counts for each gene across all cells belonging to the same cluster × sample combination. This creates one pseudobulk profile per biological replicate per cell type
pb <- aggregateData(sce,
  assay = "counts",   # use raw counts
  fun = "sum",        # sum (not mean!) — edgeR/DESeq2 expect counts
  by = c("cluster_id", "sample_id"))

pb
```

```{r aggregate-inspect}
# One assay per cell type
assayNames(pb)

# Dimensions of first cell type: genes × samples
dim(assay(pb, 1))

# Look at the pseudobulk count matrix for the first cell type
assay(pb, 1)[1:5, 1:5]
```

# Step 3: Quality control on pseudobulk

## MDS plot

```{r mds-plot, fig.width=10, fig.height=6}
# Before running DE, it is important to check whether samples separate by condition rather than by unwanted factors (batch, donor)
pbMDS(pb)
# Samples from the same condition should cluster together. If samples cluster by batch or donor instead, you may need to include covariates in the design
```
## Cell counts per group

```{r cell-counts, fig.width=10, fig.height=5}
# Visualize the number of cells contributing to each pseudobulk profile
cd <- as.data.frame(colData(sce))
raw_cd = cd %>%
  group_by(cluster_id, sample_id, group_id) %>%
  summarise(n = n(), .groups = "drop") %>%
  ggplot(aes(x = sample_id, y = n, fill = group_id)) +
  geom_col() +
  facet_wrap(~ cluster_id, scales = "free_y") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(y = "Number of cells", x = "Sample", fill = "Condition")
raw_cd
# Visualize the proportions contributing to each pseudobulk profile
pro_cd = cd %>%
  group_by(cluster_id, sample_id, group_id) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(sample_id) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = sample_id, y = prop, fill = group_id)) +
  geom_col() +
  facet_wrap(~ cluster_id, scales = "free_y") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(y = "Proportion", x = "Sample", fill = "Condition")
pro_cd
```

# Step 4: Run differential state analysis

```{r run-de}
# Runs pseudobulk DE analysis for each cell type independently. By default, it uses **edgeR** with the quasi-likelihood F-test (QLF)
res <- pbDS(pb, verbose = FALSE)
```

```{r extract-results}
# Extract tidy results table (all cell types combined)
tbl <- resDS(sce, res, bind = "row")

# gene       — gene name
# cluster_id — cell type
# logFC      — log2 fold change (stim vs ctrl)
# logCPM     — average log2 counts per million
# F          — F-statistic (edgeR QLF)
# p_val      — raw p-value
# p_adj.loc  — FDR adjusted within each cluster
# p_adj.glb  — FDR adjusted globally (across all clusters)

# Inspect the results
head(tbl)
```

```{r results-summary}
# Genes tested per cell tested
tbl %>%
  group_by(cluster_id) %>%
  summarise(
    n_tested = n(),
    n_sig_local = sum(p_adj.loc < 0.05, na.rm = TRUE),
    n_sig_global = sum(p_adj.glb < 0.05, na.rm = TRUE)
  )
```

## Filter significant genes

```{r filter-sig}
# Significant DE genes: FDR < 0.05 and |logFC| > 1
# p_adj.loc - FDR correction applied *within* each cluster. Use this if you are interested in a specific cell type.
# p_adj.glb - FDR correction applied *across all clusters*. More conservative; use this when comparing results across cell types simultaneously.

tbl_sig <- tbl %>%
  filter(p_adj.loc < 0.05, abs(logFC) > 1)

# Summary per cell type
tbl_sig %>%
  group_by(cluster_id) %>%
  summarise(
    n_up   = sum(logFC > 0),
    n_down = sum(logFC < 0),
    n_total = n()
  ) %>%
  arrange(desc(n_total))
```

# Step 5: Visualize results

## Volcano plot

```{r volcano, fig.width=7, fig.height=5}
# cell type of interest
ct <- "B cells"

df_volcano <- tbl %>%
  filter(cluster_id == ct) %>%
  mutate(
    sig = p_adj.loc < 0.05 & abs(logFC) > 1,
    label = ifelse(sig & rank(-abs(logFC)) <= 10, gene, NA)
  )

ggplot(df_volcano, aes(x = logFC, y = -log10(p_val), color = sig)) +
  geom_point(alpha = 0.5, size = 1.2) +
  scale_color_manual(values = c("grey60", "#E85D5D"), labels = c("NS", "Sig")) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "grey40") +
  geom_text_repel(aes(label = label), size = 3, max.overlaps = 15, color = "black") +
  theme_minimal(base_size = 13) +
  labs(
    title = paste0("Volcano Plot: ", ct),
    subtitle = "Stimulated vs. Control",
    x = "log2 Fold Change",
    y = "-log10(p-value)",
    color = "Significant"
  ) +
  theme(legend.position = "bottom")
```

## Heatmap of top DE genes

```{r heatmap, fig.width=10, fig.height=7}
# Built-in muscat heatmap — top 20 DE genes per cluster
pbHeatmap(sce, res, top_n = 20)
```

## Expression frequency plot

```{r freq-plot, fig.width=8, fig.height=5}
# For a specific cluster, plot the fraction of cells expressing top DE genes
top_genes <- tbl_sig %>%
  filter(cluster_id == ct) %>%
  arrange(p_adj.loc) %>%
  head(10) %>%
  pull(gene)

plotExpression(sce[, which(sce$cluster_id == ct)], #drop cells with NAs samples
  features = top_genes,
  x = "group_id",
  colour_by = "group_id") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Number of DE genes across cell types

```{r bar-de, fig.width=8, fig.height=4}
de_summary <- tbl_sig %>%
  mutate(direction = ifelse(logFC > 0, "Up", "Down")) %>%
  group_by(cluster_id, direction) %>%
  summarise(n = n(), .groups = "drop")

ggplot(de_summary, aes(x = reorder(cluster_id, n), y = n, fill = direction)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Up" = "#E85D5D", "Down" = "#4ECDC4")) +
  coord_flip() +
  theme_minimal(base_size = 13) +
  labs(
    title = "DE Genes per Cell Type",
    x = NULL,
    y = "Number of DE genes (FDR < 0.05, |logFC| > 1)",
    fill = "Direction"
  )
```

# Alternative DE methods

```{r alt-methods, eval=FALSE}
# DESeq2
res_deseq2 <- pbDS(pb, method = "DESeq2", verbose = FALSE)

# limma-voom
res_limma <- pbDS(pb, method = "limma-voom", verbose = FALSE)
```

# Adding covariates

```{r covariates, eval=FALSE}
# Example: include batch as a covariate
# Ensure batch info is in colData of the pseudobulk object
# Kang data has no batch variables
pb$batch <- metadata_table$batch[match(pb$sample_id, metadata_table$sample_id)]

# Custom design matrix
design <- model.matrix(~ 0 + group_id + batch, data = colData(pb))
contrast <- makeContrasts(group_idstim - group_idctrl, levels = design)

res_adj <- pbDS(pb, design = design, contrast = contrast, verbose = FALSE)
```

